Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-07-08 joint_inf.tex
Making progress understanding :pooh:

                    %EM Shouldn't this be \{1\}\\ ? Or perhaps the ordering of internal nodes is not what I thought?

@@

The full Farris tree log likelihood takes one of three values depending on branch lengths---assume for now the we are only interested in the ancestral state split $\{1\}$ for the site split $\{1,3\}$.

@@

(1-x'^2+y'^2-x'^2y'^2) & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
(1+x'^2-y'^2-x'^2y'^2) & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
(1+x'^2+y'^2+x'^2y'^2) & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\

@@

%EM Afraid I don't see how all of the p_emptyset terms are bounded above here.

@@

%EM I need help understanding this application of Gibbs. p_{0} + p_{13} is not 1 if I understand correctly, and so I don't see how Gibbs applies.
%You are totally right! That was an oversight on my part. I have redrawn the bounds on this term, and it will most likely affect the region of inconsistency, which is good, because I suspected something was making the inconsistency region larger than it should be. Though the bounds are not as tight as they could be now. Thanks!

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
$$
a_{1}' = a_{1}-2p_{13},
$$
$$
a_{2}' = a_{2}+2p_{13},
$$
$$
a_{3}' = a_{3}-2p_{13},
$$
$$
a_{4}' = a_{4}+2p_{13}.
$$
Then,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
%EM Sorry to be slow, but could you provide a little more detail on why this follows? Is there some convexity of entropy argument going on here?
%Not sure which part is confusing, but:
    %1) a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
    %The below equation follows by rewriting 1+x and 1-x for the above x.
    %2) a_1' and a_2' are just defined by rewriting the likelihood. There are now two fewer log(1+x) terms and two more log(1-x) terms since we replaced a (1+x)^2 in the likelihood with a (1-x)^2. That's why we have a -2p_{13} and a +2p_{13} respectively.
    %Happy to clarify somewhere in the text but not sure what part would most need exposition.
%EM Ha! Still working on understanding. It seems to me that the statement is equivalent to saying
%EM a_{1} \log a_{1} + a_{2} \log a_{2} + (1-p_{13}) \log(1-p_{13}) + p_{13}\cdot\log(2p_{13}) = a_{1}' \log a_{1}' + a_{2}' \log a_{2}'
%EM am I thinking about this wrong? Is this what you are saying in your 2)?
    %Hmm, I don't know if I see what you're saying now.
    %The way I think about it is like this: a_1 is the coefficient for the log(1+x) term and a_2 for log(1-x).
    %It is the sum of all the generating probabilities that had a log(1+x) term in it.
    %Looking at Table 2, since our first bound (we calculated it for \eta_j = {1}) was constructed with the p_{13} term of
    %   (1+x)^2(1-w)(1+y)^2,
    %our second bound (that for \eta_j = \emptyset) will be constructed with the p_{13} term of
    %   (1-x)^2(1+w)(1+y)^2.
    %So we lose 2 log(1+x)s (hence a_1' = a_1 - 2p_{13}) and gain two log(1-x)s (hence a2' = a2 + 2p_{13}).
    %Now every generating probability has log(1+w) so we replace the last two terms (p_{13}log(1-w) + (1-p_{13})log(1+w)) with just log(1+w)---maximized at log(2).
    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + \log(2) \\
    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
\end{align*}
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
    &\qquad + \log(2) \\
    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.

>>>>

@@

The full Farris tree log likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).
We will begin by writing out the middle case: for the ancestral state split $\{1\}$.

@@

1-x'^2+y'^2-x'^2y'^2 & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
1+x'^2-y'^2-x'^2y'^2 & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
1+x'^2+y'^2+x'^2y'^2 & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\

@@

@@

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
$$
a_{1}' = a_{1}-2p_{13},
$$
$$
a_{2}' = a_{2}+2p_{13},
$$
$$
a_{3}' = a_{3}-2p_{13},
$$
$$
a_{4}' = a_{4}+2p_{13}.
$$
Then, to bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + \log(2) \\
    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
    &\qquad + \log(2) \\
    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.
